---
title: Exploration of the U.S. National Oceanic and Atmospheric Administration's (NOAA)
  storm database to address questions about severe weather events
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to answer two questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The data used for this project was downloaded from: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 on 20/06/2019.

## Data Processing
```{r, echo=TRUE}
# Load required libraries
library(dplyr)
library(ggplot2)
# Load the dataset and analyse the variables
setwd("~/Documents/MOOCs/Data Specialization/5_Reproducible_Research/Project 2")
storm <- read.csv("repdata_data_StormData.csv")
names(storm)
str(storm)
# Only variables EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP will be used in the analysis.
# FATALITIES and INJURIES are variables related to harm caused to population health and PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP are related to economic consequences.
```
### Data Processing to Question 1
```{r, echo=TRUE}
# Fatalities and Injuries were aggregated by event type (EVTYPE)
stormPopHealth <- storm %>% group_by(EVTYPE) %>% summarise(TotalFatalities = sum(FATALITIES), TotalInjuries = sum(INJURIES))
# Checking total number of fatalities and injuries.
sum(stormPopHealth$TotalFatalities)
sum(stormPopHealth$TotalInjuries)
# Sort event type (EVTYPE) by most fatalities and most injuries. 
stormPopHealth2 <- stormPopHealth[order(-stormPopHealth$TotalFatalities, -stormPopHealth$TotalInjuries),]
# Analysing total number of fatalities and injuries.
head(stormPopHealth2, 10)
sum(stormPopHealth2$TotalFatalities[1:10])/sum(stormPopHealth$TotalFatalities)
sum(stormPopHealth2$TotalInjuries[1:10])/sum(stormPopHealth$TotalInjuries)
stormPopHealth3 <- stormPopHealth2[1:10,]
```
### Data Processing to Question 2
```{r, echo=TRUE}
# Checking the PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP variables
summary(storm$PROPDMGEXP)
summary(storm$CROPDMGEXP)
str(storm$PROPDMG)
str(storm$CROPDMG)
# Crop and Property Damage are expressed with numerical values while PROPDMGEXP and CROPDMGEXP are expressed with letters. Most of the letters are K, M and B which, according to NOAA documentation, mean, respectively, thousands, millions and billions.
```
A subset was created with only the variables EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP.
```{r, echo=TRUE}
stormEconomic <- subset(storm, select = c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```
Converting PROPDMGEXP and CROPDMGEXP into characters
```{r, echo=TRUE}
stormEconomic[,c(3,5)] %>% mutate_if(is.factor, as.character) -> stormEconomic2
stormEconomic$PROPDMGEXP <- stormEconomic2$PROPDMGEXP
stormEconomic$CROPDMGEXP <- stormEconomic2$CROPDMGEXP
```
Moreover, only the observations in which PROPDMGEXP or CROPDMGEXP were expressed as k, K, m, M, b, B were selected.
```{r, echo=TRUE}
stormCropEconomic <- stormEconomic[(stormEconomic$CROPDMGEXP == "k" | stormEconomic$CROPDMGEXP == "K" | stormEconomic$CROPDMGEXP == "m" | stormEconomic$CROPDMGEXP == "M"| stormEconomic$CROPDMGEXP == "B"),]
stormPropertyEconomic <- stormEconomic[(stormEconomic$PROPDMGEXP == "K" | stormEconomic$PROPDMGEXP == "m" | stormEconomic$PROPDMGEXP == "M"| stormEconomic$PROPDMGEXP == "B"),]
```
Subsetting the dataset only to include the relevant columns
```{r, echo=TRUE}
stormCropEconomic <- stormCropEconomic[,c(1,4,5)]
stormPropertyEconomic <- stormPropertyEconomic[,c(1,2,3)]
```
Substitute letters k, K, m, M, b, B in PROPDMGEXP and CROPDMGEXP for values 
```{r, echo=TRUE}
stormCropEconomic$VALUE <- as.numeric(c("B" = "1000000000", "k" = "1000", "K" = "1000", "m" = "1000000", "M" = "1000000")[stormCropEconomic$CROPDMGEXP])
stormPropertyEconomic$VALUE <- as.numeric(c("B" = "1000000000", "k" = "1000", "K" = "1000", "m" = "1000000", "M" = "1000000")[stormPropertyEconomic$PROPDMGEXP])
```
Multiply the values by PROPDMG and CROPDMG
```{r, echo=TRUE}
stormCropEconomic$TotalValue <- stormCropEconomic$CROPDMG*stormCropEconomic$VALUE
stormPropertyEconomic$TotalValue <- stormPropertyEconomic$PROPDMG*stormPropertyEconomic$VALUE
```
Aggregate by event type (EVTYPE) 
```{r, echo=TRUE}
stormCropEconomic2 <- stormCropEconomic %>% group_by(EVTYPE) %>% summarise(TotalDamagePerTypeEvent = sum(TotalValue))
stormCropEconomic2 <- stormCropEconomic %>% group_by(EVTYPE) %>% summarise(TotalDamagePerTypeEvent = sum(TotalValue))
stormPropertyEconomic2 <- stormPropertyEconomic %>% group_by(EVTYPE) %>% summarise(TotalDamagePerTypeEvent = sum(TotalValue))
```
Sort event type (EVTYPE) by highest total damage. 
```{r, echo=TRUE}
stormCropEconomic3 <- stormCropEconomic2[order(-stormCropEconomic2$TotalDamagePerTypeEvent),]
stormPropertyEconomic3 <- stormPropertyEconomic2[order(-stormPropertyEconomic2$TotalDamagePerTypeEvent),]
```
Analysing total damage cost per type of event.
```{r, echo=TRUE}
sum(stormCropEconomic3$TotalDamagePerTypeEvent[1:10])/sum(stormCropEconomic3$TotalDamagePerTypeEvent)
sum(stormPropertyEconomic3$TotalDamagePerTypeEvent[1:10])/sum(stormPropertyEconomic3$TotalDamagePerTypeEvent)
stormCropEconomic4 <- stormCropEconomic3[1:10,]
stormPropertyEconomic4 <- stormPropertyEconomic3[1:10,]
```
## Results
### Types of events most harmful with respect to population health
Total Fatalities
```{r, echo=TRUE}
op <- par(mar=c(11,4,4,2))
barplot( height= stormPopHealth3$TotalFatalities, names.arg=stormPopHealth3$EVTYPE, horiz=F, las=2,col="skyblue", main="Total Fatalities per Event Type")
rm(op)
```
Total Injuries
```{r, echo=TRUE}
op <- par(mar=c(11,4,4,2))
barplot( height= stormPopHealth3$TotalInjuries, names.arg=stormPopHealth3$EVTYPE, horiz=F, las=2,col="skyblue", main="Total Injuries per Event Type")
rm(op)
```
### Types of events with the greatest economic consequences
Total Crop Damage in US Dollars
```{r, echo=TRUE}
stormCropEconomic4
```
Total Property Damage in US Dollars
```{r, echo=TRUE}
stormPropertyEconomic4
```